---
title: "Quality control of HI titre data with hicontrol"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quality control of HI titre data with hicontrol}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
```

```{r setup}
library(hicontrol)
```

## Introduction

Haemagglutination inhibition (HI) assays measure the level of antibodies
against an influenza strain by finding the highest serum dilution that
still blocks haemagglutination. In large-scale surveillance programmes a
**reference (control) serum** of known reactivity is included in every
assay run. Because this serum should be stable over time, its titre should
remain close to a fixed target value. Sustained departures from that target
signal a laboratory process problem — such as reagent degradation, a change
in antigen quality, or an error in serum dilution — rather than genuine
biological change.

`hicontrol` applies eight statistical control chart rules to repeated HI
measurements of the same reference serum. Two properties of HI titres make
standard control chart software awkward to use:

- **Interval censoring.** Titres are reported on a doubling-dilution scale
  (10, 20, 40, 80, 160, …), so each reading is known only to the nearest
  log₂ unit. The package therefore works in log₂ units.
- **Left censoring.** Readings below the assay detection limit are reported
  as `"<10"`, `"<20"`, and so on. These are treated as half the detection
  limit before log-transformation.

---

## Data preparation: `log_num()`

Raw HI titres arrive as character strings. `log_num()` converts them to the
log₂ scale relative to a base dilution of 10, so:

| Titre  | log₂ value |
|--------|-----------|
| `"10"` | 0         |
| `"20"` | 1         |
| `"40"` | 2         |
| `"80"` | 3         |
| `"160"`| 4         |
| `"320"`| 5         |

Left-censored values are assigned half the detection limit before
transformation, so `"<10"` (treated as 5) becomes log₂(5/10) = −1.

```{r log-num}
raw <- c("80", "160", "40", "80", "160", "<10", "40", "320")
suppressWarnings(log_num(raw))
```

The warning from `as.numeric()` on the `"<10"` string is expected and
harmless; use `suppressWarnings()` in production code.

---

## A complete QC workflow

### Setting up

We create two simulated 20-run time series for a reference serum. Both use
`centre = 3` (corresponding to titre 80 on the 1:10 scale) and
`threshold = 1` (one standard deviation unit on the log₂ scale).

```{r data}
centre    <- 3
threshold <- 1
n_length  <- c(1, 3, 8, 9, 25, 10, 4, 2)  # window lengths for the 8 rules

# In-control: random scatter around centre, no violations expected
dat_ok <- c(3, 4, 2, 3, 4, 3, 2, 4, 3, 3, 4, 2, 3, 4, 3, 2, 3, 4, 3, 4)

# Out-of-control: abrupt upward step at run 11 (reagent or dilution error)
dat_ooc <- c(3, 4, 2, 3, 4, 3, 2, 4, 3, 3, 7, 7, 6, 7, 7, 6, 7, 7, 6, 7)
```

### Applying the rules: `hi_rules()`

`hi_rules()` runs all eight rules in a single call. It returns a list with
two elements:

- `[[1]]` — a list of eight raw results, one per rule. Rules that detect
  individual point violations (rules 1, 2, 3, 8) return integer vectors of
  flagged indices; rules that detect runs (rules 4–7) return matrices with
  one column per flagged run (row 1 = start index, row 2 = run length).
- `[[2]]` — a numeric vector of eight **false-positive rates**: the
  proportion of observations that were flagged by each rule.

```{r hi-rules-ok}
res_ok <- hi_rules(dat_ok, centre, threshold)

# False-positive rates — all zero for the in-control series
res_ok[[2]]
```

```{r hi-rules-ooc}
res_ooc <- hi_rules(dat_ooc, centre, threshold)

# Several rules fire after the step change
res_ooc[[2]]
```

The non-zero entries correspond to:

- **Rule 1** (1 point ≥ 3 SD): the post-step values are 4 SD above centre.
- **Rule 3** (8 consecutive points on one side): all ten post-step
  observations lie above centre.
- **Rule 4** (9 points ≥ 1 SD, either side): a relaxed version of rule 3.
- **Rule 8** (single-step difference ≥ 3): the jump between run 10 and
  run 11 is |3 − 7| = 4 log₂ units.

To inspect which runs were flagged by a specific rule:

```{r inspect-results}
# Rule 8 flagged indices (steps of >= 3 log2 units)
res_ooc[[1]][[8]]

# Rule 7 (trend) result matrix: start index, run length, direction (+1 up / -1 down)
# (returns NULL when no trend is detected)
res_ooc[[1]][[7]]
```

### Visualising with `plot_nelson()`

`plot_nelson()` draws the time series with horizontal reference lines at the
centre and at ±1, ±2, and ±3 SD, then overlays coloured points and line
segments for each rule violation. Each rule has a fixed colour:

| Colour  | Rule |
|---------|------|
| Red     | 1 — point ≥ 3 SD |
| Orange  | 2 — 2 of 3 points ≥ 2 SD |
| Yellow  | 3 — 8 points on one side |
| Tan     | 4 — 9 points ≥ 1 SD (relaxed) |
| Blue    | 5 — 25 identical values |
| Magenta | 6 — 10 alternating points |
| Pink    | 7 — monotone trend |
| Purple  | 8 — large single step |

```{r plot-nelson-ok, fig.cap = "In-control series: no rule violations."}
plot_nelson(dat_ok, res_ok[[1]], centre, threshold, n_length)
```

```{r plot-nelson-ooc, fig.cap = "Out-of-control series: abrupt step change triggers multiple rules."}
plot_nelson(dat_ooc, res_ooc[[1]], centre, threshold, n_length)
```

The large purple segment marks the rule-8 step; the red points in the
post-step region mark rule-1 violations; the tan shading connects the
rule-4 run.

---

## The eight control rules

```{r rules-table, echo = FALSE}
rules <- data.frame(
  Rule = 1:8,
  Function = c(
    "rule_xyz(x=1, y=1, z=3)",
    "rule_xyz(x=2, y=3, z=2)",
    "rule_xyz(x=8, y=8, z=0)",
    "rule_noCrelax(n=9)",
    "rule_nodiff(n=25)",
    "rule_alt(n=10)",
    "rule_trend(n=4)",
    "rule_diff(d=3)"
  ),
  Description = c(
    "1 point ≥ 3 SD from centre",
    "2 of 3 consecutive points ≥ 2 SD from centre",
    "8 consecutive points on the same side of centre",
    "9 consecutive points ≥ 1 SD from centre (either side)",
    "25 consecutive identical values",
    "10 consecutive alternating up-down points",
    "4 consecutive points in a monotone trend",
    "Single-step difference ≥ 3 log₂ units"
  ),
  Signals = c(
    "Extreme outlier; transcription or pipetting error",
    "Incipient shift or persistent mild outlier",
    "Sustained systematic bias (antigen or reagent issue)",
    "Systematic bias including zone-boundary values",
    "Frozen / stuck assay; discretisation artefact",
    "Over-correction between runs",
    "Progressive drift (reagent degradation, storage issue)",
    "Abrupt process change between consecutive runs"
  )
)
knitr::kable(rules)
```

---

## Visualisation variants

### `plot_hi()`: symmetric y-axis

`plot_hi()` produces the same chart as `plot_nelson()` but with a
symmetric y-axis centred on `centre`. This makes it immediately obvious
whether violations are above or below the centre line, and prevents a
crowded y-axis when most points cluster near the centre. Additional
arguments are passed through to `plot()`.

```{r plot-hi, fig.cap = "Symmetric-axis chart: the magnitude of the step change is easier to judge."}
plot_hi(dat_ooc, res_ooc[[1]], centre, threshold, n_length,
        main = "Reference serum — step-change example",
        ylab = expression(log[2] ~ titre))
```

### `plot_hi2()`: compact panel format

`plot_hi2()` draws a minimal chart designed for assembling many
antigen/serum pairs into a single multi-panel figure. Point and line sizes
are scaled by `ttl_len` (the total number of time points across all panels)
so that every panel appears visually consistent regardless of how many
observations it contains. When any rule fires, a pink background is drawn
as an at-a-glance alert.

```{r plot-hi2, fig.width = 4, fig.height = 3, fig.cap = "Compact panel: pink background flags at least one rule violation."}
res2_ooc <- hi_rules2(dat_ooc, centre, threshold)
plot_hi2(dat_ooc, res2_ooc[[1]], centre, threshold,
         n_length, ttl_len = 30)
```

---

## Counts instead of rates: `hi_rules2()`

`hi_rules2()` is identical to `hi_rules()` except that element `[[2]]`
contains raw counts of flagged observations rather than proportions. This
is more convenient when aggregating results across many antigen/serum
combinations, for example to rank which pairs trigger the most violations
across a surveillance season.

```{r hi-rules2}
res2_ooc <- hi_rules2(dat_ooc, centre, threshold)

# Number of observations flagged per rule
res2_ooc[[2]]

# Verify: counts / n == proportions (for rules that return vectors)
res2_ooc[[2]][c(1, 8)] / length(dat_ooc)
res_ooc[[2]][c(1, 8)]
```

The rule results list `[[1]]` is identical between `hi_rules()` and
`hi_rules2()`.

---

## Individual rule functions

All eight underlying rule functions are exported, so you can apply them
individually, combine them in a different way, or use them in custom
analysis pipelines.

```{r individual-rules}
# A 3-SD single-point outlier
dat_spike <- c(3, 4, 3, 4, 6.5, 4, 3, 4, 3, 4, 3, 4, 3, 4, 3, 4, 3, 4, 3, 4)
rule_xyz(dat_spike, centre, threshold, x = 1, y = 1, z = 3)  # index 5

# A monotone trend
dat_trend <- c(3, 3, 4, 5, 6, 7, 4, 3, 4, 3)
result <- rule_trend(dat_trend, n = 4)
result  # 3-row matrix: start, length, direction (+1 up / -1 down)

# Large single-step differences
rule_diff(dat_ooc, d = 3)  # index 10: the step at run 11
```

Every function returns `NULL` when no violations are detected, making
them safe to use in `if (!is.null(...))` guards.

---

## Reference panel charts: `ref_panel_plot()`

For large surveillance datasets stored as Acmacs map objects (from the
**Racmacs** package), `ref_panel_plot()` automates the full pipeline. It
iterates over every antigen/serum pair that has at least `min_n` repeat
measurements, applies `hi_rules2()`, and writes a multi-panel PDF of
compact `plot_hi2()` charts — one chart per qualifying pair.

```{r ref-panel, eval = FALSE}
library(Racmacs)
map <- read.acmap("my_map.ace")

ref_panel_plot(
  map,
  name   = "2024H3",
  min_n  = 5,
  file   = "qc_reference_panel_2024H3.pdf"
)
```

Each panel shows the log₂ titre series for one antigen/serum pair, with a
pink background if any rule fires. Coloured text in the bottom-right corner
of each panel lists the count for each rule; a legend in the top-right
gives the estimated standard deviation. Panels are sized so that one PDF
page fits the full matrix of qualifying pairs.
